
### Homework #5 Answers to Exercises

The text file "scrabble4letter.txt" contains over 4,000 four-letter Scrabble words. Read the .txt file into R as follows:

```{r comment="", results='hold'}
words <- scan(file="scrabble4letter.txt",what="")
```

Write regular expressions (using grep) to find the following items.

`r .exNum('All words that start with a "z"')`

`r .exNum('All words that have "zz" in them')` 

`r .exNum('All words that do not have "a", "e", "i", "o", or "u"')` 

`r .exNum('In which letter position are you most likely to find an "r"')` 

`r .exNum('Words that rhyme with "mitt"')` 


# Solutions to the exercises 
1. `r .exerciseQuestions[1]`
```{r comment=""}
grep("^[z]", words, value=TRUE)
```
2. `r .exerciseQuestions[2]`
```{r comment=""}
grep("zz", words, value=TRUE)
```
3. `r .exerciseQuestions[3]`
```{r comment=""}
grep("[^aeiou]{4}", words, value=TRUE)
```
4. `r .exerciseQuestions[4]`
```{r comment=""}
length(grep("^[r]",words,value=TRUE)) #1st
length(grep("[a-z][r][a-z]{2}",words,value=TRUE)) #2nd
length(grep("[a-z]{2}[r][a-z]",words,value=TRUE)) #3rd
length(grep("[r]$",words,value=TRUE)) #4th
```
5. `r .exerciseQuestions[5]`
```{r comment="", results='hold'}
a<- grep("itt?$", words, value=TRUE)
setdiff(a, c("suit", "toit", "doit", "gait", "wait", "bait"))
