---
title: "Introduction to NIBRS and SQL"
author:
- affiliation: University of Pennsylvania
  email: gridge@upenn.edu
  name: Greg Ridgeway
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
  html:
    theme:
      dark: [darkly, dark.scss]
      light: default
    toc: true
    html-math-method: mathjax
    embed-resources: true
  pdf:
    toc: true
    exclude: true
    prefer-html: true
number-sections: true
editor_options:
  chunk_output_type: console
---


<!-- In terminal -->
<!-- quarto render 05_Working_with_NIBRS_data.qmd -->

<!-- git commit 05_* -m "commit message" -->
<!-- git status -->
<!-- git push -->


<!-- A function for automating the numbering and wording of the exercise questions -->
```{r}
#| echo: false
# Use \x60 in place ` backtick in exercise questions
.counterExercise <- 0
.exerciseQuestions <- NULL
.exNum <- function(.questionText="")
{
   .counterExercise <<- .counterExercise+1
   .exerciseQuestions <<- c(.exerciseQuestions, .questionText)
   return(paste0(.counterExercise,". ",.questionText))
}
.exQ <- function(i)
{
   return( paste0(i,". ",.exerciseQuestions[i]) )
}
```

# National Incident-Based Reporting System

The FBI established the Uniform Crime Reporting (UCR) Program in 1930 to generate reliable crime statistics for law enforcement administration, operation, and management. It has been the primary source of crime data in the United States for decades.

The UCR's Summary Reporting System (SRS) is the best-known component of the UCR. About 18,000 law enforcement agencies, including municipal police departments, sheriff's departments, campus police, transit police, park police, and many other agencies, reported monthly counts of Part 1 crimes (murder and non-negligent manslaughter, forcible rape, robbery, aggravated assault, burglary, larceny theft, motor vehicle theft, and arson), often called "index crimes." The SRS also tracked monthly counts of Part II crimes, such as simple assault, fraud, vandalism, drug offenses, and driving under the influence. Even though reporting to the SRS was voluntary, almost all law enforcement agencies reported their data, offering fairly comprehensive coverage of crimes reported to the police in the United States.

While the UCR SRS has provided valuable crime data for many years, it has several notable limitations. 

1.  Limited number of crimes. The SRS focused on a limited number of crimes, potentially overlooking important details and emerging trends. With SRS data researchers cannot separate out trends in shootings, identity theft, and cybercrimes. Shootings, for example, are lumped together in an aggravated assault category that includes a range of serious assaults like striking someone with a beer bottle. 
2.  Only most serious crime reported. The SRS also operated under the "Hierarchy Rule," reporting only the most severe offense in a multi-offense incident.
3.  No crime details. The SRS only collected aggregate counts, meaning that detailed information about the context of the crimes, such as the characteristics of the victims and offenders, the extent of property loss or damage, the time, place, and context of the crime, and the relationships between these crime features, is lost.

The FBI introduced the National Incident-Based Reporting System (NIBRS) in the 1980s. NIBRS aimed to address the shortcomings of the SRS by capturing incident-level data and a comprehensive description of what happened in each incident. Unlike the SRS, NIBRS collects data on each individual crime incident, capturing detailed information about the offenses, victims, offenders, property, and arrestees. NIBRS records data on 52 "Group A" offenses and 10 "Group B" offenses, covering a broader spectrum of criminal activity. It did away with the Hierarchy Rule and collects data on all offenses within a single incident, providing a fuller picture of criminal activity. Because NIBRS is incident-based, we have full access to the multivariate relationships between features of crime incidents.

On January 1, 2021, the FBI officially retired the SRS marking a significant transition towards the exclusive use of the National Incident-Based Reporting System for crime data collection and reporting. Although the transition has been planned for almost a decade, many law enforcement agencies are yet to transition their information systems to report to NIBRS. In 2024, four of the nation's largest states, California, New York, Pennsylvania, and Florida, had just begun to genuinely participate in NIBRS. You can find a [map describing NIBRS coverage](https://bjs.ojp.gov/national-incident-based-reporting-system-nibrs) at the Bureau of Justice Statistics. The largest law enforcement agencies in those states regularly post crime data to their local open data portals, but you may not find those data in NIBRS yet. The transition to NIBRS also makes the study of long-term national crime trends challenging. Any study that spans the SRS-NIBRS transition will have to grapple with inconsistencies and gaps in data as agencies adapt to the new system.

The FBI accepts data from law enforcement agencies through March of the following year. That is, the FBI accepted NIBRS data for 2025 through March 2026. Some crimes committed in December 2025, for example, may be solved or result in an arrest in April 2026. Such a case would not be marked as "cleared" in the NIBRS data because the clearance came after the NIBRS 2025 submission end date. Crimes in January 2025 may seem to be solved at a higher rate than crimes in December 2025 because of the March 2026 censoring. This is a feature of the data that simply requires care. As our earlier work with the NCVS emphasized, NIBRS also contains only crimes reported to the police, and reporting rates can vary greatly by type of crime, victim features, geography, and other characteristics.

# Structured Query Language (SQL) and SQLite

The NIBRS 2025 master file contains 70 million records, far too large to load into R's memory all at once, and too complex to manage as a collection of flat files. This is where a relational database management system becomes essential. Rather than storing all data in one monolithic data frame, a relational database organizes information into multiple tables that are linked together by shared keys. In the case of NIBRS, that means separate tables for offenses, victims, offenders, property, and arrests, all linked by the ORI and incident number that uniquely identify each incident. This structure eliminates redundancy, keeps related data together, and makes it possible to answer complex multi-table questions efficiently. Perhaps most importantly for our purposes, a relational database lets you query a specific slice of the data without loading everything into memory first. A question like "how many completed burglaries occurred in Texas in 2025?" can be answered in a fraction of a second against 70 million records without R ever seeing any row outside of Texas.

Structured Query Language, or SQL (pronounced "sequel"), is the standard language for communicating with a relational database. SQL has been standardized by the American National Standards Institute since 1986 and is supported by virtually every database system in existence. One of SQL's most important characteristics is that it is a *declarative* language rather than an *imperative* one. In R you write code that tells the computer *how* to do something: loop over these rows, filter this column, group by that variable. SQL instead lets you describe *what* you want, and the database engine figures out the most efficient way to retrieve it. You write `SELECT crime_type, COUNT(*) FROM offense GROUP BY crime_type` and the engine decides on its own how to scan, filter, and aggregate the data as fast as possible. This frees you to think about the question rather than the mechanics of retrieval, and it means the same SQL query will run correctly (and fast) whether the table has a thousand rows or a billion.

We will use [SQLite](https://www.sqlite.org) as our database engine. SQLite differs from most database systems in one important way: it stores the entire database, all tables, indexes, and metadata, in a single ordinary file on disk. There is no server to install, no background process to manage, and no administrator credentials to configure. You simply point your code at the `.db` file and start querying. SQLite is also released into the public domain, meaning there are no licensing fees, no restrictions on use, and no vendor lock-in. Despite its simplicity, SQLite is far from a toy system. It is the most widely deployed database engine in the world by a large margin.  Estimates place the number of active SQLite databases in the trillions. It runs inside every iPhone and Android phone, every Mac and Windows 10 or later machine, every Chrome and Firefox browser, every Airbus A350 flight information system, and countless other devices and applications. The skills you develop querying SQLite transfer directly to PostgreSQL, MySQL, Microsoft SQL Server, Oracle, and other enterprise database systems, since the core SQL language is largely the same across all of them.

# Acquiring the data

The complete NIBRS data are available from the FBI's Crime Data Explorer [downloads page](https://cde.ucr.cjis.gov/LATEST/webapp/#/pages/downloads) under the Master File Downloads section heading. From the dropdown menu, select "National Incident-Based Reporting System (NIBRS)." The compressed data file, `nibrs-2025.zip`, is over 500 MB, so downloading will take some time. Leave it in the ZIP archive because R can read the compressed file directly.

NIBRS is too large to work with comfortably in R alone, so we will load it into a SQLite database. Once the data is in SQLite, we can query any subset of the 70 million records nearly instantaneously without ever loading the entire dataset into R's memory. We will use the `RSQLite` package to connect R to SQLite.

Let's start by loading the libraries we will need and peeking at a few lines of the data.

```{r}
#| warning: false
#| message: false
library(readr)
library(dplyr)
library(RSQLite)

# Set rebuildDB to TRUE to rebuild the database from the master file
#   (deletes any existing nibrs2025.db and re-reads all 70 million records,
#    which takes several minutes). Set to FALSE to reuse the existing
#   nibrs2025.db and skip the slow rebuild.
rebuildDB <- FALSE

# find out what files are packed inside the zip file
unzip("NIBRS/nibrs-2025.zip", list = TRUE)
# create a connection to the compressed data file
con <- unz("NIBRS/nibrs-2025.zip", "2025_NIBRS_NATIONAL_MASTER_FILE.txt")
# read first 5 lines
scan(con, nlines=5, what="", sep="\n")
```

As you can see, the data are not pretty. These data are in "fixed-width format." Rather than separating each field with a delimiter such as a comma or tab, fixed-width data place all fields side by side. This legacy format has some benefits. First, there is no need to store delimiters. This matters less today because storage is inexpensive, but delimiters in a dataset this size would require about 1 GB just to store commas. Instead, the data come with a separate file that describes which positions correspond to each field. The following table shows the first five fields of the NIBRS offense segment in the included `NIBRS Records Description updated.xlsx` file.

| Positions | Field Length and Type | Field Name                              |
|-----------|-----------------------|-----------------------------------------|
| 1-2       | A2                    | SEGMENT LEVEL                           |
| 3-4       | N2                    | NUMERIC STATE CODE                      |
| 5-13      | A9                    | ORIGINATING AGENCY IDENTIFIER (ORI)     |
| 14-25     | A12                   | INCIDENT NUMBER                         |
| 26-33     | A8                    | INCIDENT DATE                           |
| 34-36     | A3                    | UCR OFFENSE CODE                        |
| ...       | ...                   | ...                                     |

The table tells us that for offense segments, the first two characters represent the segment level, characters 3 and 4 capture the state numeric code, characters 5-13 capture the ORI (a unique identifier for a law enforcement agency), and so on. The second column describes the data type, (A)lphanumeric or (N)umeric, and the width (number of characters) of the data field.

NIBRS uses separate tables for administrative records, offenses, victims, offenders, property, and arrests. The data file interleaves records from all of these tables. The first two characters of each row indicate which table the row belongs to.

| Segment code | Segment type        |
|--------------|---------------------|
| BH           | batch header        |
| 01           | administrative      |
| 02           | offenses            |
| 03           | property            |
| 04           | victims             |
| 05           | offender            |
| 06           | arrestee            |
| 07           | Group B arrests     |
| W1           | Incomplete admin    |
| W3           | Incomplete property |
| W6           | Incomplete arrest   |

The W1, W3, and W6 codes identify "window segments" and represent partial reporting. These are relatively rare records and mostly relate to arrests or recovered property connected to offenses that do not appear in the offense segment, most likely because the agency transitioned to NIBRS between the offense and the recovery or arrest.

# The NIBRS format lookup table

I have created a lookup table called [`nibrs_format.csv`](NIBRS/nibrs_format.csv) and included it with the course materials. This table tells R exactly how wide each column is and what type of data it contains for every segment.

```{r}
nibrs_format <- read.csv("NIBRS/nibrs_format.csv")
head(nibrs_format)
```

The `nibrs_format` table has four columns:

- `col_name`: the cleaned-up column name we will use in our database
- `col_type`: "c" for character (text) or "n" for numeric
- `col_width`: the number of characters that this column occupies in the fixed-width file
- `segment`: the two-character segment code (01, 02, 03, 04, 05, 06, 07, BH)

Here, for example, are the column specifications for the offense segment.

```{r}
nibrs_format |> filter(segment=="02")
```

The code used to create `nibrs_format.csv` from the Excel documentation file is included for reference in @sec-create-nibrs-format. You do not need to run it to complete this chapter.

# Setting up the SQLite database

We will create a SQLite database called `nibrs2025.db` to store all the NIBRS segments. Once built, you can query the database from R in any future session without rebuilding it... simply reconnect.

```{r}
#| include: false
if (!rebuildDB && !file.exists("NIBRS/nibrs2025.db")) {
  stop("nibrs2025.db does not exist. Set rebuildDB <- TRUE to create it.")
}

if (rebuildDB && file.exists("NIBRS/nibrs2025.db"))
   file.remove("NIBRS/nibrs2025.db")
```

```{r}
# Create a new database (or connect if it already exists)
con <- dbConnect(SQLite(), dbname="NIBRS/nibrs2025.db")
```

# Loading the data into SQLite

The master file interleaves all the segment records, including the window-segment variants, in a single enormous file. Our goal is to separate them into eight destination tables in the SQLite database. Here is the strategy.

We will read the master file one million rows at a time. Each batch gets split by the first two characters of each line, which is the segment code. Then, for each segment in that batch, we parse the fixed-width lines directly into a data frame and append it to the appropriate SQLite table. We repeat until the file is exhausted.

The key to making this fast is `I()`. Normally `read_fwf()` expects a file path as its first argument. Wrapping a character vector in `I()`, which stands for "as-is", tells `read_fwf()` to treat the vector itself as the data rather than as a path to a file. This means we can feed each batch of raw text lines directly to `read_fwf()` without first writing them to an intermediate file. The result is that we read the master file exactly once and write each row to SQLite exactly once, rather than touching the data twice with a set of intermediate files in between.

The other performance trick here is `dbBegin()` and `dbCommit()`. By default, every `dbWriteTable()` call is its own transaction.  SQLite flushes its write buffer to disk after every one. With 70 million records spread across dozens of loop iterations, that amounts to hundreds of individual disk commits and can slow the process to a crawl. `dbBegin()` opens a single transaction that spans the entire loop. SQLite accumulates all the writes in memory and only commits them to disk once when `dbCommit()` is called at the end. This can make bulk loading an order of magnitude faster.

Building the database can take many minutes and will produce a `nibrs2025.db` database file that is over 5 GB.

```{r loadNIBRS}
#| eval: !expr rebuildDB
# Map segment codes to SQLite table names
seg_tables <- list(
 "01" = "admin",
 "02" = "offense",
 "03" = "property",
 "04" = "victim",
 "05" = "offender",
 "06" = "arrestee",
 "07" = "groupb_arrest",
 "BH" = "batch_header"
)

# Start fresh — remove any existing tables
for (tbl in unlist(seg_tables)) {
  if (dbExistsTable(con, tbl)) dbRemoveTable(con, tbl)
}

infile <- unz("NIBRS/nibrs-2025.zip", 
              "2025_NIBRS_NATIONAL_MASTER_FILE.txt",
              open = "r") # keep open for reading
dbBegin(con)  # open one transaction for the entire load
linesRead <- 0
while ((length(a <- readLines(infile, n=1000000)) > 0))
{
   linesRead <- linesRead + length(a)
   if(linesRead <= 10000000 ||       # less than 10M
      linesRead %% 10000000 == 0 ||  # a multiple of 10M
      length(a) <  1000000)          # the last lines read   
   {
      message("Lines read: ", format(linesRead, big.mark=",", scientific=FALSE))
   }
   
   dSplit <- split(a, substring(a, 1, 2))
   # Merge window segments with their corresponding complete segments
   dSplit[["01"]] <- c(dSplit[["01"]], dSplit[["W1"]])
   dSplit[["03"]] <- c(dSplit[["03"]], dSplit[["W3"]])
   dSplit[["06"]] <- c(dSplit[["06"]], dSplit[["W6"]])
   dSplit[c("W1","W3","W6")] <- NULL
   
   for (seg in names(dSplit))
   {
      fmt <- nibrs_format |> filter(segment == seg)
      
      df <-
         read_fwf(I(dSplit[[seg]]),            # parse directly from memory
                  col_positions  = fwf_widths(fmt$col_width),
                  col_type       = paste(fmt$col_type, collapse=""),
                  show_col_types = FALSE) |>
         rename_with(~fmt$col_name) |>
         rename(ori     = originating_agency_identifier_ori,
                segment = segment_level,
                state   = numeric_state_code)
      
      # Segment-specific adjustments
      if (seg == "03")
         df <- df |> mutate(value_of_property = as.numeric(value_of_property))
      
      # population covered is split by counties (for agencies spanning counties)
      #   compute total population
      if (seg == "BH")
         df <- df |> 
         mutate(total_pop = rowSums(across(starts_with("current_pop")),
                                    na.rm=TRUE))
      
      dbWriteTable(con, seg_tables[[seg]], df, append=TRUE)
   }
}

dbCommit(con)  # flush everything to disk in one shot
close(infile)
```


Let's check what tables are now in our database.

```{r}
dbListTables(con)
```

Our database now has eight data tables, one for each NIBRS segment. (We will add lookup tables later to translate the many codes in the data into readable labels.) Let's look at one record from Philadelphia in the administrative table.

```{r}
dbGetQuery(con, "
    SELECT * 
    FROM admin 
    WHERE ori='PAPEP0000' AND
          incident_number='2H-0507296XO'
    LIMIT 1")
```

The incident occurred in Pennsylvania (state 37), reported by the Philadelphia Police Department (ORI PAPEP0000), on January 6, 2025. The incident number 2H-0507296XO is only unique within an agency, so whenever linking to other segments always join on **both** `ori` and `incident_number`.

From here on, all exploration happens through SQL queries. Every time you want to work with your database in a new R session, just reconnect. No need to reload the data:

```{r}
#| eval: false
con <- dbConnect(SQLite(), dbname="NIBRS/nibrs2025.db")
```

# Database normalization

A relational database usually stores related information in several tables rather than repeating everything in one enormous table. The process of organizing the data this way is called **database normalization**. Its basic principle is to store each fact in one place and connect related facts with keys.

NIBRS is naturally organized this way. The `admin` table stores information about an incident, while the `offense`, `victim`, `offender`, `property`, and `arrestee` tables store the things associated with that incident. One incident can involve several offenses, several victims, and several offenders. If all of this information were stored in one flat table, the incident information would have to be repeated for every possible combination of offense, victim, and offender. Besides wasting space, that repetition creates the possibility of inconsistent values for what should be the same incident.

Lookup tables are another form of normalization. For example, the `offense` table stores a compact UCR offense code, while `offense_lookup` stores the readable crime description associated with each code. If the description were repeated in every offense row, a spelling change or correction would require updating millions of records. With a lookup table, the description is stored once.

The tradeoff is that answering a question often requires a `JOIN`. In NIBRS, `incident_number` is unique only within a reporting agency, so the pair `ori` and `incident_number` is the key we use to connect incident-level tables. A key does not have to identify only one row in every table: the same incident key may correctly appear many times in `offense` or `victim`, reflecting a one-to-many relationship.

SQLite can show us the structure, or **schema**, of a table with `PRAGMA table_info()`. The result lists each column's name, storage type, missing-value restriction, default value, and whether it belongs to the primary key.

```{r}
dbGetQuery(con, "PRAGMA table_info(offense)") |>
  head(10)
```

# Exploring NIBRS with SQL

Now that the data are loaded, we can ask precise questions of our database without ever loading the full dataset into memory. We will cover the key SQL clauses using NIBRS examples. SQL is not case-sensitive, but the convention is to write keywords in ALL CAPS to distinguish them from table and column names.

## `SELECT`, `FROM`, and `WHERE`

The three primary SQL clauses are `SELECT`, `FROM`, and `WHERE`. `SELECT` chooses which columns to return. `FROM` names the table where SQL will pull the data. `WHERE` filters rows. Look at the first five offense records.

```{r}
dbGetQuery(con, "
  SELECT ori, incident_number, incident_date, ucr_offense_code
  FROM   offense
  LIMIT  5")
```
The `SELECT` clause includes the four columns that I wish to pull from the table listed in the `FROM` clause, offense. The SQL keyword `LIMIT` is like the R `head()` function and will just show us the first 5 rows of the result rather than the entire offense table.  Because there is no `WHERE` clause, this query will pull all the rows from the offense table.

We can use `*` to select all columns.
```{r}
dbGetQuery(con, "
  SELECT *
  FROM   offense
  LIMIT  3")
```

Let's look at the offense record for the Philadelphia incident we saw in the administrative segment. The `WHERE` clause can combine multiple conditions with `AND` and `OR`.

```{r}
dbGetQuery(con, "
  SELECT *
  FROM   offense
  WHERE ori='PAPEP0000' AND
        incident_number='2H-0507296XO'")
```

The UCR offense code is 120 (robbery). `offense_attempted_completed` is `C`, meaning it was a completed crime rather than an attempted crime. The offender was not suspected of using drugs, alcohol, or computer equipment (`suspected_of_using_1` = `N`), and the location type is `20` (Residence/Home).

SQL uses a single `=` to test for equality, unlike R, which uses `==`. `ori = 'PAPEP0000'` tests whether `ori` equals that value. SQLite also accepts `==` as an extension, but standard SQL uses `=`, so using a single equals sign makes queries portable to other database systems. SQL has no separate assignment operator in a `WHERE` clause, so there is no ambiguity.

Text values in SQL `WHERE` clauses go inside single quotes. In standard SQL, double quotes instead identify table or column names that contain spaces, punctuation, or SQL keywords.

Column names should ideally contain only letters, numbers, and underscores and should not be SQL keywords. When an inherited database has an awkward name containing a space, punctuation, or a keyword, the identifier must be quoted. Standard SQL uses double quotes around identifiers, but SQLite also accepts square brackets, which are convenient inside an R string. Here we deliberately give a result column an awkward alias:

```{r}
dbGetQuery(con, "
  SELECT incident_date AS [incident date]
  FROM   offense
  LIMIT  5")
```

The same brackets protect an awkward name when it appears in `SELECT`, `FROM`, `WHERE`, or another clause. For example, an existing column named `incident date` would be written as `[incident date]`.

::: {.callout-note title="Exercises"}
`r .exNum('Using the \x60victim\x60 table, show the ORI, incident number, age, and sex for the first 10 individual victims. Individual victims have \x60type_of_victim = \'I\'\x60.')`

`r .exNum('Using the \x60arrestee\x60 table, show the incident number, arrest date, age, and arrest offense code for the first 10 arrest records from the Philadelphia Police Department (ORI PAPEP0000).')`
:::

## Regular expressions in SQL

SQLite can also use regular expressions in a `WHERE` clause. Once per R session, initialize RSQLite's regular-expression extension. The following pattern selects ORIs that begin with `PA`:

```{r}
initExtension(con, "regexp")

dbGetQuery(con, "
  SELECT DISTINCT ori
  FROM   batch_header
  WHERE  ori REGEXP '^PA'
  LIMIT  10")
```


## `COUNT()`, `GROUP BY`, `ORDER BY`, and `DISTINCT`

How many offenses were reported to the Philadelphia Police Department (ORI PAPEP0000)?

```{r}
dbGetQuery(con, "
  SELECT COUNT(*) AS offense_count
  FROM   offense
  WHERE  ori='PAPEP0000'")
```

`COUNT(*)` is an aggregate function that counts rows. Unlike `COUNT(column)`, it does not skip a row because a particular column contains `NULL`. You will commonly see `COUNT()` used in three ways.

-  `COUNT(*)` for total number of rows
-  `COUNT(column)` for nonmissing values
-  `COUNT(DISTINCT column)` for unique nonmissing values

We use `AS` to give the result a readable column name.

`GROUP BY` groups rows by one or more columns and applies an aggregate function to each group. SQL aggregate functions include `COUNT()`, `SUM()`, `AVG()`, `MIN()`, and `MAX()`. `ORDER BY` then sorts the results. Add `DESC` after a column or expression to sort from largest to smallest; ascending order is the default. These clauses commonly appear together: first SQL forms the groups and calculates their counts, and then it sorts those counts.

The rule to remember is: every column in `SELECT` that is not inside an aggregate function must appear in `GROUP BY`.

Let's count offenses by UCR code to find the most commonly reported crime types.

```{r}
dbGetQuery(con, "
  SELECT ucr_offense_code,
         COUNT(*) AS offense_count
  FROM   offense
  GROUP BY ucr_offense_code
  ORDER BY offense_count DESC
  LIMIT 10")
```

Why is `ucr_offense_code` in the `GROUP BY` clause? Because it is not being aggregated in the `SELECT` clause.

::: {.callout-important}

## What goes in the `GROUP BY` clause?

_Everything_ in the `SELECT` clause not being aggregated.

What happens if I forget? You will not get an error or warning. SQL will still give you a result, but the SQL standard does not define what that result will be. Most likely it will not be correct.

:::

Those UCR codes are hard to interpret. This is a recurring pattern in NIBRS.  Many fields are stored as short codes that require a separate lookup table to decode. We will deal with this properly in a later section by loading lookup tables into SQLite and joining them. For now, let's keep exploring.

Let's count offenses by state.
```{r}
dbGetQuery(con, "
  SELECT state,
         COUNT(*) AS offense_count
  FROM   offense
  GROUP BY state
  ORDER BY offense_count DESC
  LIMIT 10")
```
Again we have codes rather than names. I made a `nibrs_state_lookup.csv` file that maps numeric state codes to names. We can load it into R temporarily to decode the top result.
```{r}
a <- read.csv("NIBRS/nibrs_state_lookup.csv")
head(a)
a |> filter(state_code == 42)
```
State 42 is Texas, which tends to report more offenses than other states, not necessarily because it has more crime, but because it has very good NIBRS participation across both large and small agencies. Shortly we will load all the lookup tables into SQLite so we can decode codes directly in SQL queries rather than bouncing back to R.

Sometimes we want the unique values rather than a count for each value. `DISTINCT` removes duplicate rows from a result, and `ORDER BY` can sort the resulting unique values. Let's find the set of offense types present in the database.
```{r}
dbGetQuery(con, "
  SELECT DISTINCT ucr_offense_code
  FROM   offense
  ORDER BY ucr_offense_code")
```
Every offense code in that list corresponds to a crime type in `nibrs_offense_lookup.csv`. Peek at it to get a sense of what is there.
```{r}
read.csv("NIBRS/nibrs_offense_lookup.csv") |> head(10)
```

Let's explore the property table.
```{r}
dbGetQuery(con, "
  SELECT *
  FROM   property
  WHERE ori='PAPEP0000' AND
        incident_number='2H-0507296XO'")
```
These show two records related to the Philadelphia robbery. With the same ORI and incident number, these relate to the same incident.  I made a lookup table for the codes in `property_description`.
```{r}
read.csv("NIBRS/nibrs_property_lookup.csv") |>
   filter(property_code == 16)
```
We are getting a clearer picture. Someone broke into someone's home while they were there and had some household goods stolen. The `type_property_loss` column explains what exactly happened.

| Code | Description |
|:-----|:------------|
| 1 | None |
| 2 | Burned (includes damage caused in fighting the fire) |
| 3 | Counterfeited/Forged |
| 4 | Destroyed/Damaged/Vandalized |
| 5 | Recovered (to impound property that was previously stolen) |
| 6 | Seized (to impound property that was not previously stolen) |
| 7 | Stolen/Etc. (includes bribed, defrauded, embezzled, extorted, ransomed, robbed, etc.) |
| 8 | Unknown |

So the household goods were stolen (Code 7) and recovered (Code 5) on the same day. The data also give the estimated value of the items, here \$20, but these values can look a little strange, so let's explore those a little more.

`MIN()`, `MAX()`, `AVG()`, and `SUM()` work with numeric columns. Here is the range of reported property values by type of property stolen. The `property_description` column is yet another code field. Code 75, for example, is "Portable Electronic Communications" (i.e., a mobile phone).
```{r}
dbGetQuery(con, "
  SELECT property_description,
         COUNT(*)                         AS n,
         MIN(value_of_property)           AS min_value,
         MAX(value_of_property)           AS max_value,
         ROUND(AVG(value_of_property), 0) AS avg_value
  FROM   property
  WHERE  value_of_property > 1        -- 1 means 'unknown value'
    AND  type_property_loss = '7'     -- 7 means stolen
  GROUP BY property_description
  ORDER BY avg_value DESC
  LIMIT 10")
```
SQL comments start with `--`. Anything after `--` on a line is ignored by the database engine.

The NIBRS documentation says "If the value of the property is unknown, the value will be one ("1") dollar." However, it is very common, particularly in older or official data collections to use 9, 99, -9, or other variations of lots of 9s to indicate missing values. Let's see if there are ORIs that seem to have a lot of 9s.
```{r}
dbGetQuery(con, "
  SELECT ori,
         value_of_property,
         COUNT(*) AS property_records
  FROM property
  WHERE value_of_property IN (
    999999999, 99999999, 9999999, 999999, 99999,
    9999, 999, 99)
  GROUP BY ori,value_of_property
  ORDER BY property_records DESC
  LIMIT 10
")
```
There are many values composed entirely of 9s, and they appear concentrated in a few departments that use this convention repeatedly.

It is hard to tell what to do with these. For now, let's replace 1 and 999999999 with `NULL`, SQL's missing value indicator.
```{r}
dbExecute(con, "
  UPDATE property
  SET    value_of_property = NULL
  WHERE  value_of_property IN (1, 999999999)")
```
We use `dbExecute()` instead of `dbGetQuery()` since we are not retrieving data from the database. `dbExecute()` will return the number of rows that the update affects.

::: {.callout-note title="Exercises"}
`r .exNum('How many completed robbery offense records (UCR code 120) were reported to the Philadelphia Police Department (ORI PAPEP0000)?')`

`r .exNum('How many distinct agencies reported at least one offense in Pennsylvania? Use \x60REGEXP\x60 to identify ORIs beginning with PA.')`

`r .exNum('How many offense records meet our definition of a shooting: aggravated assault (13A), completed (C), with firearm code 11, 12, 13, 14, or 15 in any weapon_force column?')`
:::

## Creating and changing tables

So far, `SELECT` queries have returned results to R. Sometimes we instead want to save a query result as a new table in the database. `CREATE TABLE ... AS SELECT ...` creates the table and fills it with the rows returned by the query. The following example makes a small table containing 100 randomly selected homicide records. Note the use of `RANDOM()` to put the results in a random order before taking the first 100.

```{r}
# in case they already exist
dbExecute(con, "DROP TABLE IF EXISTS homicide_sample")
dbExecute(con, "DROP TABLE IF EXISTS homicide_review_sample")

dbExecute(con, "
  CREATE TABLE homicide_sample AS
  SELECT ori, incident_number, incident_date
  FROM   offense
  WHERE  ucr_offense_code = '09A'
  ORDER BY RANDOM()
  LIMIT  100")
```

The new table is independent of the query that created it, meaning later changes to `offense` do not affect `homicide_sample`. Also, `CREATE TABLE AS` copies the query results but not indexes or constraints from the original table.

`ALTER TABLE` changes the structure of an existing table. Common operations include renaming a table, adding a column, renaming a column, and dropping a column. Let's rename that sample of homicide offenses.
```{r}
dbListTables(con)
dbExecute(con, "ALTER TABLE homicide_sample RENAME TO homicide_review_sample")
dbListTables(con)
```
The `batch_header` table contains a column called `date_state_began_ibr_for_law_enforcement_officers_killed_and_assaulted_leoka_data`, which is much too long.
```{r}
# LIKE allows partial matching
#   % means zero or more characters
#   _ means exactly one character
# pragma_table_info() gives a table of table info
dbGetQuery(con, "
  SELECT *
  FROM pragma_table_info('batch_header')
  WHERE name LIKE '%leoka%'")
```

```r
dbExecute(con, "
  ALTER TABLE batch_header RENAME COLUMN
    date_state_began_ibr_for_law_enforcement_officers_killed_and_assaulted_leoka_data
    TO date_leoka")
```

```{r}
#| include: false
invisible(try(dbExecute(con, "
  ALTER TABLE batch_header RENAME COLUMN
    date_state_began_ibr_for_law_enforcement_officers_killed_and_assaulted_leoka_data
    TO date_leoka"), silent=TRUE))
```

```{r}
dbGetQuery(con, "
  SELECT *
  FROM pragma_table_info('batch_header')
  WHERE name LIKE '%leoka%'")
```
Some agencies, such as Oklahoma City OK, Kansas City MO, and High Point NC, span multiple counties. NIBRS records their population separately by county. When creating `batch_header`, we computed `total_pop` by summing the individual county populations. Since we no longer need the individual county populations, we can remove those columns to simplify the table. Dropping columns makes their storage available for SQLite to reuse but does not necessarily make the database file smaller. A later section explains how `VACUUM` can shrink the file.
```{r}
dbGetQuery(con, "
  SELECT *
  FROM pragma_table_info('batch_header')
  WHERE name REGEXP '_[1-5]$'")
```

```r
dbExecute(con, "ALTER TABLE batch_header DROP COLUMN current_population_1")
dbExecute(con, "ALTER TABLE batch_header DROP COLUMN current_population_2")
dbExecute(con, "ALTER TABLE batch_header DROP COLUMN current_population_3")
dbExecute(con, "ALTER TABLE batch_header DROP COLUMN current_population_4")
dbExecute(con, "ALTER TABLE batch_header DROP COLUMN current_population_5")
```

```{r}
#| include: false
for (column in paste0("current_population_", 1:5)) {
  invisible(try(dbExecute(con, paste(
    "ALTER TABLE batch_header DROP COLUMN", column
  )), silent=TRUE))
}
```

```{r}
dbGetQuery(con, "
  SELECT *
  FROM pragma_table_info('batch_header')
  WHERE name REGEXP '_[1-5]$'")
```
The `batch_header` table already has `number_of_months_reported`, but let's double-check it against the month-by-month reporting fields.
```{r}
# drop the column if it already exists
if ("reported_all_12_months" %in% dbListFields(con, "batch_header")) {
  dbExecute(con, "
    ALTER TABLE batch_header
    DROP COLUMN reported_all_12_months")
}

# create the new column filled with 0s
dbExecute(con, "
  ALTER TABLE batch_header
  ADD COLUMN reported_all_12_months INTEGER NOT NULL DEFAULT 0
")

dbExecute(con, "
  UPDATE batch_header
  SET reported_all_12_months =
    CASE
      WHEN january   != 'NNN'
       AND february  != 'NNN'
       AND march     != 'NNN'
       AND april     != 'NNN'
       AND may       != 'NNN'
       AND june      != 'NNN'
       AND july      != 'NNN'
       AND august    != 'NNN'
       AND september != 'NNN'
       AND october   != 'NNN'
       AND november  != 'NNN'
       AND december  != 'NNN'
      THEN 1
      ELSE 0
    END
")

dbGetQuery(con, "
  SELECT reported_all_12_months,
         number_of_months_reported,
         COUNT(*) AS n
  FROM batch_header
  GROUP BY reported_all_12_months,
           number_of_months_reported")
```
Note that `number_of_months_reported` has several strange values. Our own calculation may be preferable.

Deleting rows or dropping tables does not necessarily reduce the size of a SQLite database file. SQLite normally keeps the freed pages available for later use. `VACUUM` rebuilds the database and returns that unused space to the operating system.

```{r}
#| eval: false
dbExecute(con, "VACUUM")
```

On a multi-gigabyte database, `VACUUM` can take a long time, requires additional free disk space while it rebuilds the file, and prevents other connections from writing to the database. It is maintenance to run when substantial space needs to be reclaimed, not part of an ordinary analysis. It also cannot run inside an active transaction.

::: {.callout-note title="Exercise"}
`r .exNum('Which 10 states have the largest number of fully reporting agencies? Use the \x60batch_header\x60 table and the \x60reported_all_12_months\x60 column.')`
:::


## Joining tables

So far, when we needed to interpret a code, we loaded the corresponding lookup file into R and searched it for a matching description. That works for an occasional code, but repeatedly moving between SQL results and R lookup tables is tedious and error-prone. A better approach is to load the lookup tables into SQLite and let SQL combine the codes with readable labels such as crime, location, weapon, property, city, and state names.

Joins are also necessary because NIBRS distributes the information for an incident across separate tables for offenses, victims, offenders, property, and arrestees. We link these incident tables using `ori` and `incident_number`. We link lookup tables using the code field and its corresponding lookup key. Both tasks use a SQL `JOIN`, so joins are central to working with NIBRS.

A `JOIN` is how SQL links two tables together. You name both tables in the `FROM` clause, then use an `ON` clause to specify which columns in each table contain the matching keys. For every row in the left table, SQL finds the matching row(s) in the right table and merges the columns together.

### Loading lookup tables into SQLite

Although `dbWriteTable()` could create and populate the lookup table in a single line, it would infer only the columns’ storage types and would not designate `ucr_code` as the table’s primary key. Creating the table explicitly requires more code, but it records an important rule in the database schema: every row must have a nonmissing, unique UCR code. SQLite will therefore reject missing or duplicated codes rather than silently creating an ambiguous lookup table. The primary-key declaration also documents how the table should be joined to the offense data and automatically creates an index that can make code lookups more efficient. After establishing this structure, `dbAppendTable()` adds the CSV data without replacing the constraints defined for the table.
```{r}
#| label: loadOffenseLookup

dbExecute(con, "DROP TABLE IF EXISTS offense_lookup")

a <- read.csv("NIBRS/nibrs_offense_lookup.csv", colClasses = "character")
head(a)

# create the table structure
#   set ucr_code to be the primary key
#     NOT NULL: every row must have a ucr_code
#     PRIMARY KEY: must be unique, no duplicates
dbExecute(con, "
  CREATE TABLE offense_lookup (
    ucr_code  TEXT NOT NULL PRIMARY KEY,
    crime_cat TEXT,
    crime     TEXT)")
# populate with data
dbAppendTable(con, "offense_lookup", a)

# note that ucr_code is now marked as the primary key
dbGetQuery(con, "PRAGMA table_info(offense_lookup)")
# can also see that SQLite has built an index for this table
dbGetQuery(con, "PRAGMA index_list(offense_lookup)")
```
We will talk more about the purpose of an index later. Let's create the remaining lookup tables.

```{r}
dbExecute(con, "DROP TABLE IF EXISTS state_lookup")
dbExecute(con, "DROP TABLE IF EXISTS location_lookup")
dbExecute(con, "DROP TABLE IF EXISTS weapon_lookup")
dbExecute(con, "DROP TABLE IF EXISTS property_lookup")
dbExecute(con, "DROP TABLE IF EXISTS drug_lookup")
dbExecute(con, "DROP TABLE IF EXISTS circumstance_lookup")

a <- read.csv("NIBRS/nibrs_state_lookup.csv", colClasses = "character")
# index_list() will not show an index for integer primary keys (it's okay)
dbExecute(con, "
  CREATE TABLE state_lookup (
    state_code INTEGER NOT NULL PRIMARY KEY,
    state_abbr TEXT,
    state_name TEXT)")
dbAppendTable(con, "state_lookup", a)

a <- read.csv("NIBRS/nibrs_location_lookup.csv", colClasses = "character")
dbExecute(con, "
  CREATE TABLE location_lookup (
    location_code TEXT NOT NULL PRIMARY KEY,
    location_type TEXT)")
dbAppendTable(con, "location_lookup", a)

a <- read.csv("NIBRS/nibrs_weapon_lookup.csv", colClasses = "character")
dbExecute(con, "
  CREATE TABLE weapon_lookup (
    weapon_code TEXT NOT NULL PRIMARY KEY,
    weapon_type TEXT)")
dbAppendTable(con, "weapon_lookup", a)

a <- read.csv("NIBRS/nibrs_property_lookup.csv", colClasses = "character")
dbExecute(con, "
  CREATE TABLE property_lookup (
    property_code TEXT NOT NULL PRIMARY KEY,
    property_type TEXT)")
dbAppendTable(con, "property_lookup", a)

a <- read.csv("NIBRS/nibrs_drug_lookup.csv", colClasses = "character")
dbExecute(con, "
  CREATE TABLE drug_lookup (
    drug_code TEXT NOT NULL PRIMARY KEY,
    drug_type TEXT)")
dbAppendTable(con, "drug_lookup", a)

a <- read.csv("NIBRS/nibrs_circumstance_lookup.csv", colClasses = "character")
dbExecute(con, "
  CREATE TABLE circumstance_lookup (
    circumstance_code TEXT NOT NULL PRIMARY KEY,
    circumstance      TEXT)")
dbAppendTable(con, "circumstance_lookup", a)
```

```{r}
dbListTables(con)
```

Our database now has sixteen tables: the eight segment tables we loaded from the master file, the seven lookup tables, and the homicide sample.

### `INNER JOIN`

An `INNER JOIN` returns only rows where the join key exists in **both** tables. Rows with no match in the other table are dropped entirely.

The syntax is:

```sql
FROM left_table
  INNER JOIN right_table
    ON left_table.key_column = right_table.key_column
```

The offense table just has offense codes. We can look at the most common crime codes reported to the Philadelphia PD.
```{r}
dbGetQuery(con, "
  SELECT ucr_offense_code,
         COUNT(*) AS offense_count
  FROM   offense
  WHERE  offense.ori = 'PAPEP0000'
  GROUP BY ucr_offense_code
  ORDER BY offense_count DESC
  LIMIT 10")
```
It's not very useful without knowing what those codes are. Let's use `offense_lookup` to replace UCR codes with readable crime names for the top offenses reported to the Philadelphia Police Department.
```{r}
dbGetQuery(con, "
  SELECT offense_lookup.crime,
         COUNT(*) AS offense_count
  FROM   offense
    INNER JOIN offense_lookup
      ON offense.ucr_offense_code = offense_lookup.ucr_code
  WHERE  offense.ori = 'PAPEP0000'
  GROUP BY offense_lookup.crime
  ORDER BY offense_count DESC
  LIMIT 10")
```
When a column name could exist in multiple tables (like `ori`), prefix it with the table name, `offense.ori`, so SQL knows exactly where to look. The general practice is to always prefix in JOIN queries for clarity, even when it is technically unambiguous.

Now let's join the offense table to `batch_header` to find the cities that report the most offenses. The `batch_header` has one row per agency and the `offense` table has one row per offense. The join brings city and state names into the offense counts.
```{r}
dbGetQuery(con, "
  SELECT batch_header.city_name,
         batch_header.state_abbreviation,
         COUNT(*) AS offense_count
  FROM   offense
    INNER JOIN batch_header
      ON offense.ori = batch_header.ori
  GROUP BY batch_header.city_name, batch_header.state_abbreviation
  ORDER BY offense_count DESC
  LIMIT 10")
```


Where do crimes happen? Joining `location_lookup` answers that immediately.
```{r}
dbGetQuery(con, "
  SELECT location_lookup.location_type,
         COUNT(*) AS offense_count
  FROM   offense
    INNER JOIN location_lookup
      ON offense.location_type = location_lookup.location_code
  GROUP BY location_lookup.location_type
  ORDER BY offense_count DESC
  LIMIT 10")
```
Residences and homes are the most frequently reported offense locations. These counts do not by themselves measure individual risk because they do not account for how much time people spend in each type of location.

What weapons are most commonly involved in aggravated assaults? The offense table stores up to three weapon codes per offense in `weapon_force_1`, `weapon_force_2`, and `weapon_force_3`. `UNION ALL` stacks all three columns into one so we can count across them together, then we join `weapon_lookup` for readable labels.
```{r}
dbGetQuery(con, "
  WITH all_weapons AS (
      SELECT weapon_force_1 AS weapon_code 
      FROM offense 
      WHERE ucr_offense_code = '13A' AND 
            weapon_force_1 IS NOT NULL
    UNION ALL
      SELECT weapon_force_2 
      FROM offense 
      WHERE ucr_offense_code = '13A' AND 
            weapon_force_2 IS NOT NULL
    UNION ALL
      SELECT weapon_force_3 
      FROM offense 
      WHERE ucr_offense_code = '13A' AND 
            weapon_force_3 IS NOT NULL
  )
  SELECT weapon_lookup.weapon_type,
         COUNT(*) AS n
  FROM   all_weapons
    INNER JOIN weapon_lookup ON 
       all_weapons.weapon_code = weapon_lookup.weapon_code
  GROUP BY weapon_lookup.weapon_type
  ORDER BY n DESC")
```
Note the use of `WITH` here. It creates a temporary result called `all_weapons` that the `FROM` clause in the main query can use. This is an example of a Common Table Expression (CTE), which we will discuss more later.

Now let's revisit the property value query from earlier. Instead of raw numeric codes we join `property_lookup` to get readable property type names.
```{r}
dbGetQuery(con, "
  SELECT property_lookup.property_type,
         COUNT(*)                         AS n,
         MIN(value_of_property)           AS min_value,
         MAX(value_of_property)           AS max_value,
         ROUND(AVG(value_of_property), 0) AS avg_value,
         MEDIAN(value_of_property)        AS med_value
  FROM   property
    INNER JOIN property_lookup
      ON property.property_description = property_lookup.property_code
  WHERE  value_of_property IS NOT NULL
    AND  type_property_loss = '7'       -- stolen
  GROUP BY property_lookup.property_type
  ORDER BY med_value DESC
  LIMIT 10")
```

::: {.callout-important title="Extreme values warrant scrutiny even after cleaning"}
Even after replacing the 1 and 999999999 with `NULL`, extreme outliers can still hide data entry errors. Here we see that there is a report of a stolen automobile worth \$729M. It seems a Mobile, Alabama car dealership got swindled out of a \$729M car on November 1, 2025. Curiously, there are two other cases of individuals losing \$729M from a parking lot/garage and from their home. Plausible?

The NIBRS data contains no quality check preventing the entering of any dollar amount. When computing aggregate property values, consider using the median, capping or excluding implausible outliers, and always inspect the top of the distribution before reporting results.
:::

To obtain more details about these incidents, I needed to join several tables. The query joins

-  `property` to `property_lookup` to identify the type of property
-  `offense` and `offense_lookup` to identify the associated crime
-  `location_lookup` to describe where the offense occurred
-  `victim` to obtain the victim type
-  `batch_header` to identify the reporting city

The `offense` and `victim` tables are linked by the combination of `ori` and `incident_number`, while the lookup tables are linked by their respective code fields.

::: {.callout-note title="Incident-level joins can produce combinations"}
The data do not directly link a particular property record to a particular victim. This query therefore joins them at the incident level. If an incident has multiple offenses and multiple victims, the result will contain every offense-by-victim combination. For example, an incident with two offenses and three victims can produce six rows for the same property record. These rows describe the records associated with the incident; they do not establish which victim was associated with each offense or item of property.
:::

Let's time the query. Because it joins several large tables, it might take several minutes.
```{r}
#| include: false
# Remove the demonstration indexes so that the first timing measures the
# query without them, even when reusing a database from an earlier render.
dbExecute(con, "DROP INDEX IF EXISTS idx_offense")
dbExecute(con, "DROP INDEX IF EXISTS idx_victim")
dbExecute(con, "DROP INDEX IF EXISTS idx_batch_header")
```

```{r}
#| label: lookupPropertyCases729M
property729_query <- "
  SELECT batch_header.city_name,
         property.ori,
         property.incident_number,
         property.incident_date,
         property_lookup.property_type,
         offense_lookup.crime,
         location_lookup.location_type,
         victim.type_of_victim
  FROM property
    INNER JOIN property_lookup
      ON property.property_description = property_lookup.property_code
    INNER JOIN offense
      ON property.ori = offense.ori
         AND property.incident_number = offense.incident_number
    INNER JOIN offense_lookup
      ON offense.ucr_offense_code = offense_lookup.ucr_code
    INNER JOIN location_lookup
      ON offense.location_type = location_lookup.location_code
    INNER JOIN victim
      ON property.ori = victim.ori
         AND property.incident_number = victim.incident_number
    INNER JOIN batch_header
      ON property.ori = batch_header.ori
  WHERE property.value_of_property = 729000000
    AND property.type_property_loss = '7'
 "

# about 4 minutes
system.time(
  property729 <- dbGetQuery(con, property729_query)
)

property729
```

### SQL indexes

Although the filter finds only a few property records, SQLite still needs to locate matching rows in the much larger `offense` and `victim` tables. Without an appropriate index, it scans millions of rows looking for a match. That is not an efficient way to find records.

SQL offers you the option of building an index. SQL indexes are just like the ones you would use in a book. Look in the index, which is sorted in a meaningful way, and locate the page that contains the topic you wanted to find. Without suitable indexes, SQL may repeatedly scan many rows to match each `ori` and `incident_number`. Composite indexes on `(ori, incident_number)` are especially useful here because those two columns together identify the incident being joined. The primary keys on the lookup tables already provide indexes for their code columns, so we do not need to create additional lookup-table indexes. You only need to build the index once. Once built, it will stay with the database, even if you restart your computer.

Let's add indexes that support the slow joins. The composite indexes on `offense` and `victim` help SQLite jump directly to the matching incident, and the `batch_header` index speeds up the agency lookup. The initial filter on `property` is already fast in this example, so it does not need another index.

```{r}
#| label: indexPropertyCases729M
dbExecute(con, "
  CREATE INDEX IF NOT EXISTS idx_offense
  ON offense(ori, incident_number)")

dbExecute(con, "
  CREATE INDEX IF NOT EXISTS idx_victim
  ON victim(ori, incident_number)")

dbExecute(con, "
  CREATE INDEX IF NOT EXISTS idx_batch_header
  ON batch_header(ori)")
```

Now rerun exactly the same query. SQLite automatically chooses an index when its query planner determines that using one will be helpful.

```{r}
#| label: lookupPropertyCases729MIndexed
system.time(
  property729_indexed <- dbGetQuery(con, property729_query)
)

property729_indexed
```
Wow! So much faster. It takes some upfront computing time to create an index (maybe a minute) and the index takes space on the computer (the offense and victim indexes each require about 400 Mb... the `batch_header` index is 0.5 Mb).

::: {.callout-note title="Exercises"}
`r .exNum('What are the five most common location types for robbery (UCR code 120)? Join \x60offense\x60 to \x60location_lookup\x60 and report the number of robbery offense records for each location type.')`

`r .exNum('How many cases are recorded as justifiable homicides? Check both the \x60offense\x60 table (UCR code 09C) and the \x60victim\x60 table (circumstance codes 20 and 21). Use \x60circumstance_lookup\x60 to report the circumstance descriptions.')`

| Code | Description |
|:-----|:------------|
| 20   | Criminal killed by private citizen |
| 21   | Criminal killed by police officer |
:::


### `LEFT JOIN`

An `INNER JOIN` drops any row from the left table that has no match in the right table. A `LEFT JOIN` keeps all rows from the left table and fills in `NULL` for any columns from the right table that did not have a match. In our work, `LEFT JOIN` is almost always the right choice. For example, we have offenses that we want to link to arrestees. If for a particular offense no arrest has been made (yet), then typically we do not want to drop the offense from our dataset. Instead, we just want an `NA` for arrestee to indicate that no arrest has been made. A left join is the right choice when you do not want to accidentally drop records.

Start by examining the arrestee table.
```{r}
dbGetQuery(con, "SELECT * from arrestee LIMIT 5")
```
These show records of five arrests that Fairbanks officers made. The data include age, sex, race, and ethnicity of the arrestee. Four of these arrests were "on-view" (`type_of_arrest` = O) and one was taken into custody based on a warrant or prior report (`type_of_arrest` = T). A third option would be a summons or citation (`type_of_arrest` = S).

There can be multiple arrests associated with the same incident. Here is a 2025 incident in Alabama in which three offenses occurred.
```{r}
dbGetQuery(con, "
  SELECT ori, incident_number, ucr_offense_code, 
         offense_lookup.crime
  FROM offense 
    LEFT JOIN offense_lookup
      ON offense.ucr_offense_code = offense_lookup.ucr_code
  WHERE ori='AL0010000' AND incident_number='E88U3T5U0O-3'")
```
The report lists three offenders.
```{r}
dbGetQuery(con, "
  SELECT ori, incident_number, offender_sequence_number,
         age_of_offender, race_of_offender, ethnicity_of_offender,
         sex_of_offender
  FROM offender
  WHERE ori='AL0010000' AND incident_number='E88U3T5U0O-3'")
```
And all three were arrested. We can line up the offender and arrestee records in this case using age, race, and sex, but offender sequence numbers do not correspond to arrestee sequence numbers.
```{r}
dbGetQuery(con, "
  SELECT ori, incident_number, arrestee_sequence_number, 
         ucr_arrest_offense_code, 
         age_of_arrestee, race_of_arrestee, 
         ethnicity_of_arrestee, sex_of_arrestee 
  FROM arrestee 
  WHERE ori='AL0010000' AND incident_number='E88U3T5U0O-3'")
```

Now that we have some familiarity with the arrestee table, let's compute the percentage of completed burglary offense records (UCR code 220) whose incident has at least one arrest record. Because we will link many offense records to the arrestee table, it is best to create an arrestee index too.
```{r}
dbExecute(con, "
  CREATE INDEX IF NOT EXISTS idx_arrestee
  ON arrestee(ori, incident_number)")
```
Not all incidents will have a matching row in the `arrestee` table, so we use a `LEFT JOIN` to keep all burglaries whether or not an arrest was made.
```{r}
dbGetQuery(con, "
  SELECT offense.incident_date,
         offense.ori,
         offense.incident_number,
         arr.ori                  AS arrest_ori,
         arr.incident_number      AS arrest_incident_number
  FROM   offense
    LEFT JOIN (SELECT DISTINCT ori, incident_number
               FROM   arrestee) AS arr
      ON offense.ori = arr.ori
         AND offense.incident_number = arr.incident_number
  WHERE  offense.ucr_offense_code = '220'
    AND  offense.offense_attempted_completed = 'C'
  LIMIT 10")
```
Where the arrest ORI and incident numbers are `NA`, the left join could not find an ORI/incident number match over in the `arrestee` table.

One offense can result in multiple arrests. Note the "subquery" used inside the left join. If there are multiple arrests for an incident, this subquery reduces them to a single row with `DISTINCT`. Otherwise, if one burglary resulted in 10 arrests, then that incident would get counted 10 times. As written, each offense gets one row and will check whether there are _any_ associated arrests.

A cleaner and more efficient route to check for the existence of a match is to use `EXISTS()`.
```{r}
dbGetQuery(con, "
  SELECT offense.incident_date,
         offense.ori,
         offense.incident_number,
         EXISTS (SELECT 1
                 FROM arrestee
                 WHERE arrestee.ori = offense.ori AND 
                       arrestee.incident_number = offense.incident_number) 
               AS arrest_made
  FROM offense
  WHERE offense.ucr_offense_code = '220'
    AND offense.offense_attempted_completed = 'C'
  LIMIT 10
")
```
Now let's count them. SQL evaluates all expressions in a `SELECT` clause simultaneously, so an alias defined in one column cannot be referenced by another column in the same `SELECT`. We cannot calculate a percentage using `total_burglaries` and `burglaries_with_arrest` in the same `SELECT` that defines those aliases; SQL does not yet know what they are. A Common Table Expression (CTE) solves this by computing the counts first and dividing in a second step.
```{r}
dbGetQuery(con, "
  WITH burglary_counts AS (
    SELECT COUNT(*) AS total_burglaries,
           SUM(CASE WHEN EXISTS (
                   SELECT 1
                   FROM arrestee
                   WHERE arrestee.ori = offense.ori
                     AND arrestee.incident_number = offense.incident_number)
                 THEN 1
                 ELSE 0
               END) AS burglaries_with_arrest
    FROM offense
    WHERE offense.ucr_offense_code = '220'
      AND offense.offense_attempted_completed = 'C')
  SELECT total_burglaries,
         burglaries_with_arrest,
         ROUND(100.0 * burglaries_with_arrest /
                       NULLIF(total_burglaries, 0), 1) AS pct_arrested
  FROM burglary_counts")
```
`CASE WHEN ... THEN ... ELSE ... END` is SQL's version of an if-else expression. Here, we add 1 when the incident has at least one arrest record and otherwise add 0. Summing those indicators gives the number of burglary offense records associated with an arrest, not the number of individual arrests. Dividing by the total number of burglary offense records gives the percentage associated with an arrest.

::: {.callout-note title="Exercise"}
`r .exNum('Among completed motor vehicle theft offense records (UCR code 240), how many have an incident with at least one arrest record, and what percentage is that? Use \x60EXISTS()\x60.')`
:::

## Date functions

SQLite stores dates as plain text. NIBRS incident dates arrive in `YYYYMMDD` format, without hyphens. Let's confirm:

```{r}
dbGetQuery(con, "
  SELECT incident_date
  FROM   offense
  ORDER BY RANDOM()
  LIMIT 10")
```

SQLite's date functions (`STRFTIME()`, `DATE()`, etc.) expect the standard `YYYY-MM-DD` format, so we need to reformat `incident_date` before we can use them. The `UPDATE` statement edits rows in place. We use `SUBSTR()` to pull apart the year, month, and day and reassemble them with hyphens. The `||` operator concatenates strings in SQL.

```{r}
dbExecute(con, "
  UPDATE offense
  SET    incident_date = SUBSTR(incident_date, 1, 4) || '-' ||
                         SUBSTR(incident_date, 5, 2) || '-' ||
                         SUBSTR(incident_date, 7, 2)
  WHERE  incident_date NOT LIKE '____-__-__'")
```

The `WHERE incident_date NOT LIKE '____-__-__'` clause is a safeguard.  Each `_` matches any single character, so the pattern matches dates already in `YYYY-MM-DD` format. This means re-running the chunk will not corrupt already-formatted dates. The same fix applies to `incident_date` and `arrest_date` in the `arrestee` table.

```{r}
dbExecute(con, "
  UPDATE arrestee
  SET    incident_date = SUBSTR(incident_date, 1, 4) || '-' ||
                         SUBSTR(incident_date, 5, 2) || '-' ||
                         SUBSTR(incident_date, 7, 2)
  WHERE  incident_date NOT LIKE '____-__-__'")

dbExecute(con, "
  UPDATE arrestee
  SET    arrest_date = SUBSTR(arrest_date, 1, 4) || '-' ||
                       SUBSTR(arrest_date, 5, 2) || '-' ||
                       SUBSTR(arrest_date, 7, 2)
  WHERE  arrest_date NOT LIKE '____-__-__'")
```

Now `STRFTIME()` works directly. It takes a format string and a date column. Common format codes are `%Y` (4-digit year), `%m` (month 01–12), `%d` (day), `%H` (hour on a 24-hour clock), and `%w` (day of week, 0=Sunday).

Importantly, `STRFTIME()` always returns **text**, even when the result looks like a number. This is fine for labels such as year-month, but convert the result with `CAST(... AS INTEGER)` before making a numerical comparison or calculation. Here we extract the month as an integer and keep offenses from October through December:

```{r}
dbGetQuery(con, "
  SELECT incident_date,
         CAST(STRFTIME('%m', incident_date) AS INTEGER) AS month
  FROM   offense
  WHERE  CAST(STRFTIME('%m', incident_date) AS INTEGER) >= 10
  LIMIT  10")
```

Let's count offenses by month to look for seasonal patterns.

```{r}
dbGetQuery(con, "
  SELECT STRFTIME('%Y-%m', incident_date) AS year_month,
         COUNT(*)                         AS offense_count
  FROM   offense
  GROUP BY year_month
  ORDER BY year_month")
```

Now let's look at whether the percentage associated with an arrest varies by the month in which the offense occurred. Remember that the March 2026 submission cutoff means crimes committed late in 2025 have less time to produce a recorded arrest.

```{r}
dbGetQuery(con, "
  WITH clearance AS (
    SELECT STRFTIME('%m', incident_date) AS month,
           ucr_offense_code,
           COUNT(*) AS total,
           SUM(CASE WHEN EXISTS (
                   SELECT 1
                   FROM arrestee
                   WHERE arrestee.ori = offense.ori
                     AND arrestee.incident_number = offense.incident_number)
                 THEN 1
                 ELSE 0
               END) AS arrests
    FROM offense
    WHERE ucr_offense_code IN ('09A','220') -- homicide and burglary
    GROUP BY month, ucr_offense_code)
  SELECT month,
         ucr_offense_code,
         total,
         arrests,
         ROUND(100.0 * arrests / NULLIF(total, 0), 1) AS pct_arrested
  FROM clearance
  ORDER BY ucr_offense_code, month")
```

The percentage associated with an arrest drops for crimes committed late in the year, not necessarily because those crimes are less likely to be solved, but because there is less time between the crime and the NIBRS submission cutoff for an arrest to be recorded.

The March 2026 submission cutoff makes raw arrest percentages hard to compare. A homicide in January has over a year for an arrest to appear in the data, while a homicide in December has only a few months. To put every crime on equal footing, we can ask a more precise question: was there an arrest *within 60 days* of the offense? Every offense in the year has a full 60-day observation window.

To compute the time between two dates, SQLite provides `JULIANDAY()`, which converts a date to the number of days since a fixed reference point (noon on November 24, 4714 BC, the start of the "Julian day" used by astronomers). Subtracting one Julian day from another gives the number of days between two dates. Let's look at the time from offense to first arrest for a few homicides (UCR code 09A, murder and non-negligent manslaughter). Because the `arrestee` table can hold several arrests per incident, we first collapse it to the earliest arrest date per incident with `MIN(arrest_date)`.

```{r}
dbGetQuery(con, "
  SELECT offense.ori,
         offense.incident_number,
         offense.incident_date,
         arr.first_arrest,
         JULIANDAY(arr.first_arrest) - 
           JULIANDAY(offense.incident_date) AS days_to_arrest
  FROM offense
    INNER JOIN (SELECT ori, incident_number, MIN(arrest_date) AS first_arrest
                FROM arrestee
                GROUP BY ori, incident_number) AS arr
      ON offense.ori = arr.ori
         AND offense.incident_number = arr.incident_number
  WHERE  offense.ucr_offense_code = '09A'
  LIMIT 10")
```

Now let's compute the homicide clearance rate by state, where "cleared" means an arrest occurred within 60 days of the offense. We use a `LEFT JOIN` so that homicides with no arrest at all still count in the denominator, and we join `state_lookup` to label the states. Note that `days_to_arrest` is `NULL` whenever there was no arrest, and any comparison with `NULL` (like `NULL <= 60`) is not `TRUE`, so those incidents correctly fall outside the cleared count.

```{r}
dbGetQuery(con, "
  WITH homicide_clearance AS (
    SELECT offense.state,
           JULIANDAY(arr.first_arrest) - 
             JULIANDAY(offense.incident_date) AS days_to_arrest
    FROM offense
      LEFT JOIN (SELECT ori, incident_number, MIN(arrest_date) AS first_arrest
                 FROM   arrestee
                 GROUP BY ori, incident_number) AS arr
        ON offense.ori = arr.ori
           AND offense.incident_number = arr.incident_number
    WHERE  offense.ucr_offense_code = '09A'
  )
  SELECT state_lookup.state_name,
         COUNT(*) AS homicides,
         SUM(CASE WHEN days_to_arrest <= 60
                  THEN 1 ELSE 0 END) AS cleared_60,
         ROUND(100.0 * SUM(CASE WHEN days_to_arrest <= 60
                                THEN 1 ELSE 0 END) / COUNT(*), 1) 
           AS pct_cleared_60
  FROM homicide_clearance
    INNER JOIN state_lookup
      ON homicide_clearance.state = state_lookup.state_code
  GROUP BY state_lookup.state_name
  HAVING homicides >= 50
  ORDER BY pct_cleared_60 DESC")
```

The `HAVING` clause is new here. It filters *groups* after aggregation, whereas `WHERE` filters individual rows *before* aggregation. The distinction matters because of when each clause is evaluated. SQL processes a query in stages: it first uses `WHERE` to decide which rows to keep, then `GROUP BY` collapses those rows into groups and computes the aggregates (`COUNT(*)`, `SUM(...)`, and so on), and only then does `HAVING` get to filter based on those computed aggregates. At the moment `WHERE` runs, the group counts do not exist yet, so a condition like `homicides >= 50` could not possibly be evaluated there.

That ordering gives us a simple rule of thumb: if your filter condition involves an aggregate function, it must go in `HAVING` and if it involves only raw column values, it goes in `WHERE`. A condition like `offense.ucr_offense_code = '09A'` looks only at raw column values, so it belongs in `WHERE` (and we put it there, inside the CTE). A condition like `COUNT(*) >= 50` depends on an aggregate, so it must go in `HAVING`. As a bonus, filtering in `WHERE` is also faster whenever it is possible, because it discards rows early, before the database does the work of grouping and aggregating them, so prefer `WHERE` for any condition that does not require an aggregate.

Here we use `HAVING homicides >= 50` to keep only states with at least 50 homicides so that the percentages are not dominated by states with just a handful of cases. The resulting 60-day clearance rates are more comparable across states than raw cleared counts because every homicide had the same 60-day observation window.

::: {.callout-note title="Exercise"}
`r .exNum('In which month were the most robberies (UCR code 120) reported? Use \x60STRFTIME()\x60 to extract the month, then group and order the results.')`
:::


## Working `WITH` CTEs

Let's use a CTE to find the states with the largest populations not yet covered by NIBRS. We will compute total population by state, split by whether the agency reports to NIBRS, then calculate what percentage of each state's population is covered.

One important SQL gotcha: agencies that do not report to NIBRS have `agency_nibrs_flag = NULL`. In SQL, any comparison with `NULL` (including `NULL != 'A'`) evaluates to `NULL`, not `TRUE`. So the condition `agency_nibrs_flag != 'A'` would silently exclude all non-reporting agencies rather than selecting them. The correct test is `agency_nibrs_flag IS NULL`.

```{r}
dbGetQuery(con, "
  WITH  -- tabulate state population by state x NIBRS reporting
    pop_by_state AS (
      SELECT state_abbreviation,
             agency_nibrs_flag,
             SUM(total_pop) AS pop
      FROM   batch_header
      WHERE  covered_by_ori IS NULL   -- exclude those reporting through other
      GROUP BY state_abbreviation, agency_nibrs_flag)
  SELECT 
    state_abbreviation,
    MAX(CASE WHEN agency_nibrs_flag='A' THEN pop ELSE 0 END) 
      AS in_nibrs,
    MAX(CASE WHEN agency_nibrs_flag IS NULL THEN pop ELSE 0 END)  
      AS not_in_nibrs,
    ROUND(100.0 * MAX(CASE WHEN agency_nibrs_flag='A' THEN pop ELSE 0 END) /
            (MAX(CASE WHEN agency_nibrs_flag='A' THEN pop ELSE 0 END) +
             MAX(CASE WHEN agency_nibrs_flag IS NULL THEN pop ELSE 0 END)), 0)
      AS pct_covered
  FROM pop_by_state
  GROUP BY state_abbreviation
  ORDER BY pct_covered")
```

The states with the lowest NIBRS coverage tend to include Florida, Pennsylvania, California, and New York, all very populous states. This is an important limitation to keep in mind when interpreting national statistics from NIBRS.

You can use multiple CTEs in one `WITH` clause, separating them with commas. The query below uses two CTEs to identify the agency with the most shootings (aggravated assault, UCR 13A, completed, firearm involved), then joins in the agency name from the batch header.

```{r}
dbGetQuery(con, "
  WITH  -- tabulate shootings by ORI
    shootings AS (
      SELECT ori, COUNT(*) AS shooting_count
      FROM   offense
      WHERE  ucr_offense_code = '13A'
        AND  offense_attempted_completed = 'C'
        AND  (weapon_force_1 IN ('11','12','13','14','15') OR
              weapon_force_2 IN ('11','12','13','14','15') OR
              weapon_force_3 IN ('11','12','13','14','15'))
      GROUP BY ori
    ),
    top_agency AS ( -- which ORI reports the most shootings?
      SELECT ori, shooting_count
      FROM   shootings
      ORDER BY shooting_count DESC
      LIMIT 1
    )
  SELECT batch_header.city_name,
         batch_header.state_abbreviation,
         top_agency.shooting_count
  FROM   top_agency
    INNER JOIN batch_header 
      ON top_agency.ori = batch_header.ori")
```

::: {.callout-note title="Exercise"}
`r .exNum('What is the race distribution of shooting victims? Use a CTE to identify shooting incidents, join to \x60victim\x60, and search all ten victim UCR offense-code fields to confirm that each victim was connected to aggravated assault (13A).')`
:::

## Window functions

So far we have not really worked with the `victim` table much. Let's peek at the first few rows.

```{r}
dbGetQuery(con, "
  SELECT *
  FROM   victim
  LIMIT  5")
```

Each row is one victim connected to one incident. `type_of_victim` distinguishes an individual person (`I`) from a business (`B`), government (`G`), society (`S`), or other victim types. Many NIBRS offenses (shoplifting, vandalism, fraud, drug offenses) have a business or society as the "victim" rather than a person. 

| Code | Type of victim          |
|:-----|:------------------------|
| I    | Individual              |
| B    | Business                |
| F    | Financial Institution   |
| G    | Government              |
| R    | Religious Organization  |
| S    | Society/Public          |
| L    | Law Enforcement Officer |
| O    | Other                   |
| U    | Unknown                 |

For individual victims, NIBRS records `age_of_victim`, `sex_of_victim` (`M`/`F`), `race_of_victim`, and `ethnicity_of_victim` as separate fields. The `ucr_offense_code_1` (through `_10`) columns record which offenses in the incident this particular victim was connected to, since a single incident can have several offenses and several victims who are not all tied to every offense.

A natural question is how victim demographics differ across crime types. We already know how to count victims by sex within each crime using `GROUP BY`. But we often want each count expressed as a *percentage of that crime's victims*, and that requires dividing each group's count by the total for its crime. With the tools we have so far, that means computing the per-crime totals in one query and joining them back. A **window function** does it in a single step.

A window function computes an aggregate across a set of related rows but, unlike `GROUP BY`, it does not collapse those rows. Instead it attaches the computed value to every row. You write an ordinary aggregate followed by an `OVER` clause that defines the "window" of rows to aggregate over. The simplest window is `OVER ()`, meaning "all rows in the result." Here we compute each race's share of all individual victims. We use a CTE to get the per-race counts first, then a window function to divide by the grand total.

```{r}
dbGetQuery(con, "
  WITH victim_race AS (
    SELECT race_of_victim,
           COUNT(*) AS n
    FROM   victim
    WHERE  type_of_victim = 'I'      -- individuals only
    GROUP BY race_of_victim)
  SELECT race_of_victim,
         n,
         ROUND(100.0 * n / SUM(n) OVER (), 1) AS pct_of_victims
  FROM   victim_race
  ORDER BY n DESC")
```

`SUM(n) OVER ()` adds up `n` across every row of the result and reports that single grand total alongside each row, so dividing gives each race's percentage. Without window functions we would have needed a second query (or a self-join) just to get that total. Normally you would see the aggregate function `SUM()` in the `SELECT` clause and automatically start thinking that you need a `GROUP BY` to make this work. Here, however, we do not want the query to collapse across any group. The `OVER ()` clause communicates how we want `SUM()` to aggregate, and it simply attaches that total to every row.

The real power comes from `PARTITION BY`, which splits the rows into groups and restarts the window aggregate within each group, much like `GROUP BY` but without collapsing the rows. The query below breaks down victim sex *within* each crime type. The `PARTITION BY offense_lookup.crime` makes `SUM(COUNT(*))` total only the victims of that one crime, so the percentages sum to 100 within each crime. By searching all ten offense-code fields with `IN`, the query counts each victim–crime association. A victim associated with multiple crimes will, therefore, contribute once to the count for each crime.

```{r}
a <- dbGetQuery(con, "
  SELECT offense_lookup.crime,
         victim.sex_of_victim AS sex,
         COUNT(*) AS n,
         ROUND(100.0 * COUNT(*) /
               SUM(COUNT(*)) OVER (PARTITION BY offense_lookup.crime), 1) 
           AS pct_within_crime
  FROM victim
    INNER JOIN offense_lookup
      ON offense_lookup.ucr_code IN (
        victim.ucr_offense_code_1, victim.ucr_offense_code_2,
        victim.ucr_offense_code_3, victim.ucr_offense_code_4,
        victim.ucr_offense_code_5, victim.ucr_offense_code_6,
        victim.ucr_offense_code_7, victim.ucr_offense_code_8,
        victim.ucr_offense_code_9, victim.ucr_offense_code_10)
  WHERE  victim.type_of_victim = 'I'
    AND  victim.sex_of_victim IN ('M','F')
  GROUP BY offense_lookup.crime, victim.sex_of_victim
  ORDER BY offense_lookup.crime, victim.sex_of_victim")
a
```

Notice the nested `SUM(COUNT(*))`: the inner `COUNT(*)` is the ordinary aggregate produced by the `GROUP BY` (the number of victims of a given crime and sex), and the outer `SUM(...) OVER (PARTITION BY ...)` adds those group counts together within each crime to give the crime's total.

Standard SQL has no built-in "pivot" option, but we can do that in R to get a different view of these results.
```{r}
library(tidyr)
a |>
   pivot_wider(id_cols = crime,
               names_from = sex,
               values_from = pct_within_crime) |>
   arrange(desc(F)) |> 
   # first 6 and last 6
   slice(1:6, (n()-5):n())
```

The results reveal stark differences in who is victimized by different crimes, with offenses such as forcible rape skewing heavily female while robbery skews male.

::: {.callout-note title="Exercise"}
`r .exNum('Which 10 law enforcement agencies account for the largest shares of all arrest records? Count arrest records by ORI, then use a window function to calculate each agency\'s percentage of the national total.')`
:::

When you are done and ready to shutdown, be sure to disconnect from your database.
```{r}
#| eval: false
dbDisconnect(con)
```

# Answers to the exercises

`r .exQ(1)`
```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT ori, incident_number, age_of_victim, sex_of_victim
  FROM   victim
  WHERE  type_of_victim = 'I'
  LIMIT  10")
```

`r .exQ(2)`
```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT incident_number, arrest_date, age_of_arrestee,
         ucr_arrest_offense_code
  FROM   arrestee
  WHERE  ori = 'PAPEP0000'
  LIMIT  10")
```

`r .exQ(3)`
```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT COUNT(*) AS offense_count
  FROM   offense
  WHERE  ori = 'PAPEP0000'
    AND  ucr_offense_code = '120'
    AND  offense_attempted_completed = 'C'")
```

`r .exQ(4)`
```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT COUNT(DISTINCT ori) AS pa_agency_count
  FROM   offense
  WHERE  ori REGEXP '^PA'")
```

`r .exQ(5)`
```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT COUNT(*) AS shooting_count
  FROM   offense
  WHERE  ucr_offense_code = '13A'
    AND  offense_attempted_completed = 'C'
    AND  (weapon_force_1 IN ('11','12','13','14','15') OR
          weapon_force_2 IN ('11','12','13','14','15') OR
          weapon_force_3 IN ('11','12','13','14','15'))")
```

`r .exQ(6)`
```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT state_abbreviation,
         COUNT(*) AS fully_reporting_agencies
  FROM   batch_header
  WHERE  reported_all_12_months = 1
  GROUP BY state_abbreviation
  ORDER BY fully_reporting_agencies DESC
  LIMIT 10")
```

`r .exQ(7)`

```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT location_lookup.location_type,
         COUNT(*) AS robbery_count
  FROM   offense
    INNER JOIN location_lookup
      ON offense.location_type = location_lookup.location_code
  WHERE  offense.ucr_offense_code = '120'
  GROUP BY location_lookup.location_type
  ORDER BY robbery_count DESC
  LIMIT 5")
```

`r .exQ(8)`

Using the offense table:
```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT COUNT(*) AS justifiable_homicide_count
  FROM   offense
  WHERE  ucr_offense_code = '09C'")
```

Using the victim table, joining `circumstance_lookup` to replace both circumstance-code fields with descriptions:
```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT circumstance_lookup.circumstance,
         COUNT(*) AS victim_count
  FROM   victim
    INNER JOIN circumstance_lookup
      ON circumstance_lookup.circumstance_code IN (
           victim.circumstance_1, victim.circumstance_2)
  WHERE  circumstance_lookup.circumstance_code IN ('20','21')
  GROUP BY circumstance_lookup.circumstance")
```

`r .exQ(9)`
```{r}
#| comment: ""
dbGetQuery(con, "
  WITH theft_counts AS (
    SELECT COUNT(*) AS total_thefts,
           SUM(CASE WHEN EXISTS (
                 SELECT 1
                 FROM arrestee
                 WHERE arrestee.ori = offense.ori
                   AND arrestee.incident_number = offense.incident_number)
               THEN 1 ELSE 0 END) AS thefts_with_arrest
    FROM offense
    WHERE ucr_offense_code = '240'
      AND offense_attempted_completed = 'C'
  )
  SELECT total_thefts,
         thefts_with_arrest,
         ROUND(100.0 * thefts_with_arrest /
               NULLIF(total_thefts, 0), 1) AS pct_with_arrest
  FROM theft_counts")
```

`r .exQ(10)`
```{r}
#| comment: ""
dbGetQuery(con, "
  SELECT STRFTIME('%m', incident_date) AS month,
         COUNT(*) AS robbery_count
  FROM   offense
  WHERE  ucr_offense_code = '120'
  GROUP BY month
  ORDER BY robbery_count DESC
  LIMIT 1")
```

`r .exQ(11)`
```{r}
#| comment: ""
dbGetQuery(con, "
  WITH shooting_incidents AS (
    SELECT DISTINCT ori, incident_number
    FROM   offense
    WHERE  ucr_offense_code = '13A'
      AND  offense_attempted_completed = 'C'
      AND  (weapon_force_1 IN ('11','12','13','14','15') OR
            weapon_force_2 IN ('11','12','13','14','15') OR
            weapon_force_3 IN ('11','12','13','14','15'))
  )
  SELECT victim.race_of_victim,
         COUNT(*) AS victim_count
  FROM   shooting_incidents
    INNER JOIN victim
      ON shooting_incidents.ori = victim.ori
         AND shooting_incidents.incident_number = victim.incident_number
  WHERE '13A' IN (
    victim.ucr_offense_code_1, victim.ucr_offense_code_2,
    victim.ucr_offense_code_3, victim.ucr_offense_code_4,
    victim.ucr_offense_code_5, victim.ucr_offense_code_6,
    victim.ucr_offense_code_7, victim.ucr_offense_code_8,
    victim.ucr_offense_code_9, victim.ucr_offense_code_10)
  GROUP BY victim.race_of_victim
  ORDER BY victim_count DESC")
```

`r .exQ(12)`
```{r}
#| comment: ""
dbGetQuery(con, "
  WITH agency_arrests AS (
    SELECT ori,
           COUNT(*) AS arrests
    FROM   arrestee
    GROUP BY ori
  )
  SELECT ori,
         arrests,
         ROUND(100.0 * arrests / SUM(arrests) OVER (), 2)
           AS pct_of_all_arrests
  FROM   agency_arrests
  ORDER BY arrests DESC
  LIMIT 10")
```

```{r}
#| echo: false
dbDisconnect(con)
```

# Creating `nibrs_format.csv` {#sec-create-nibrs-format .appendix}

The following code was used to create `nibrs_format.csv` from the Excel documentation file. It is included here for reference and does not need to be run.

```{r}
#| eval: false
library(readxl)
library(tidyr)

fmtXL <- read_excel("NIBRS Records Description updated.xlsx",
                    sheet = "INCIDENT RECORD",
                    range = "A5:D819") |>
   rename(DataField=`Data Field Number`,
          TypeLength=`Type/ Length`)

# Helper to pull format rows for one segment and clean them up
clean_fmt <- function(rows, exclude_pos=NULL) {
  fmt <- fmtXL |>
    slice(rows) |>
    filter(!is.na(Position))
  if (!is.null(exclude_pos))
    fmt <- fmt |> filter(!(Position %in% exclude_pos))
  data.frame(
    col_name  = fmt$Description |>
      strsplit(split=" - ", fixed=TRUE) |>
      sapply(head, n=1) |>
      gsub("[^A-Za-z0-9]+", "_", x=_) |>
      gsub("_+$", "", x=_) |>
      tolower(),
    col_type  = fmt$TypeLength |>
      substring(1,1) |>
      recode_values(from=c("A","N"), to=c("c","n")),
    col_width = fmt$TypeLength |>
      substring(2) |>
      as.numeric()
  )
}

# Locate the start of each segment in fmtXL
i02 <- grep('LEVEL "02"', fmtXL$DataField)
i03 <- grep('LEVEL "03"', fmtXL$DataField)
i04 <- grep('LEVEL "04"', fmtXL$DataField)
i05 <- grep('LEVEL "05"', fmtXL$DataField)
i06 <- grep('LEVEL "06"', fmtXL$DataField)
i07 <- grep('LEVEL "07"', fmtXL$DataField)

nibrs_format <- bind_rows(
  clean_fmt(3:(i02-3),    "59-88")                              |> mutate(segment="01"),
  clean_fmt((i02+1):(i03-3), c("38-40","46-48","49-57"))       |> mutate(segment="02"),
  clean_fmt((i03+2):(i04-3), c("14-22","20-22","58-102","58-72","103-132")) |> mutate(segment="03"),
  clean_fmt((i04+1):(i05-3), c("37-66","74-77","79-83","84-123")) |> mutate(segment="04"),
  clean_fmt((i05+1):(i06-3))                                   |> mutate(segment="05"),
  clean_fmt((i06+2):(i07-3), c("61-66","75-104"))              |> mutate(segment="06"),
  clean_fmt(-(1:i07),     "44-49")                             |> mutate(segment="07")
)

# The property segment has two columns both named ESTIMATED.QUANTITY;
# the second represents thousandths of the first, so rename it.
nibrs_format <- nibrs_format |>
  mutate(col_name = if_else(
    segment=="03" &
    col_name=="estimated_quantity" &
    duplicated(paste(segment, col_name)),
    "estimated_quantity_1000ths",
    col_name))

# Shorten the unwieldy type_property_loss_etc name
nibrs_format <- nibrs_format |>
  mutate(col_name = if_else(col_name=="type_property_loss_etc",
                            "type_property_loss",
                            col_name))

# Batch header lives on a different sheet
fmtBH <- read_excel("NIBRS Records Description updated.xlsx", skip=4) |>
  data.frame() |>
  filter(!is.na(Position) &
         !(Position %in% c("106-225","234-269","234","235","236","270-284")))

nibrs_format <- bind_rows(
  nibrs_format,
  data.frame(
    col_name  = fmtBH$Description |>
      strsplit(split=" - ", fixed=TRUE) |>
      sapply(head, n=1) |>
      gsub("[^A-Za-z0-9]+", "_", x=_) |>
      gsub("_+$", "", x=_) |>
      tolower(),
    col_type  = fmtBH$Type..Length |>
      substring(1,1) |>
      recode_values(from=c("A","N"), to=c("c","n")),
    col_width = fmtBH$Type..Length |>
      substring(2) |>
      as.numeric(),
    segment   = "BH"
  )
)

write_csv(nibrs_format, "NIBRS/nibrs_format.csv")
```
